---
title: "Weighted spatial tessellations in constrained domains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighted spatial tessellations in constrained domains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup}
library(sf)
library(terra)
library(weightedVoronoi)
```

# Introduction

Spatial allocation around point locations is a recurring requirement in ecology and social–ecological research, including the delineation of settlement influence zones, species territories, and management areas. Voronoi tessellations provide a simple and transparent framework for dividing space among generators, but standard implementations assume uniform influence and unconstrained Euclidean distance.

The weightedVoronoi package extends this framework by allowing (i) generator-specific weights and (ii) tessellations constrained to arbitrary polygonal domains using either Euclidean or geodesic distance. This vignette demonstrates the use of the package on a representative example with a complex domain boundary.


## Example data

We first define a concave, irregular spatial domain and a set of generator points with heterogeneous weights.

```r
crs_use <- 32636  # projected CRS (metres)

make_irregular_boundary <- function(crs = crs_use) {
  # A simple concave polygon (fast, deterministic, vignette-safe)
  coords <- rbind(
    c(0, 0),
    c(1200, 0),
    c(1200, 900),
    c(800, 900),
    c(800, 500),
    c(400, 500),
    c(400, 900),
    c(0, 900),
    c(0, 0)
  )
  st_sf(geometry = st_sfc(st_polygon(list(coords))), crs = crs)
}

boundary_sf <- make_irregular_boundary()

set.seed(42)
pts <- st_sample(boundary_sf, size = 8, type = "random")

points_sf <- st_sf(
  village = paste0("V", seq_along(pts)),
  population = round(exp(rnorm(length(pts), log(300), 0.8))),
  geometry = pts,
  crs = crs_use
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Weighted Euclidean tessellation

We first generate a weighted tessellation using Euclidean distance. Generator influence is scaled using a logarithmic transformation of the population weights.

```r
out_euc <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "euclidean",
  # Optional alternative weight semantics:
  weight_model = "multiplicative",
  clip_to_boundary = TRUE,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_euc$polygons["generator_id"], add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Weighted geodesic tessellation

Geodesic tessellations compute shortest-path distances constrained to the domain.

```r
out_geo <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "geodesic",
  close_mask = TRUE,
  close_iters = 1,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_geo$polygons["generator_id"], add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```
## Custom resistance surfaces and barriers

In many ecological and social–ecological applications, effective distance is shaped not only by domain geometry but also by environmental resistance. Topography, land cover, and infrastructure can increase or block movement.

The recommended workflow is to build a resistance surface (optionally composed from multiple rasters), apply barriers, and then pass the result to weighted_voronoi_domain() via resistance_rast.

```r
# Build a template grid (aligned to the analysis)
template <- out_euc$allocation

# Two simple resistance layers (for demonstration)
R_land <- template; terra::values(R_land) <- 1
R_land[template == 1] <- 2  # small spatial pattern

R_risk <- template; terra::values(R_risk) <- 1
xy <- terra::xyFromCell(R_risk, 1:terra::ncell(R_risk))
band <- xy[,1] > 450 & xy[,1] < 550
vals <- terra::values(R_risk); vals[band] <- 10; terra::values(R_risk) <- vals

# Compose layers (multiplicative by default)
R <- compose_resistance(R_land, R_risk, template = template, method = "multiply")

# Add a river barrier (semi-permeable)
river <- st_sf(
  geometry = st_sfc(st_linestring(rbind(c(500, 0), c(500, 900)))),
  crs = st_crs(boundary_sf)
)

R2 <- add_barriers(R, river, permeability = "semi", cost_multiplier = 20, width = 20)

out_geo_res <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "geodesic",
  resistance_rast = R2,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_geo_res$polygons["generator_id"], add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Geodesic tessellations with elevation-dependent resistance

In many ecological and social–ecological applications, effective distance is
shaped not only by domain geometry but also by environmental resistance.
Topography, land cover, or infrastructure may increase the cost of movement
between locations. To accommodate this, `weightedVoronoi` allows geodesic
distance calculations to be modified by an isotropic resistance surface.

When a digital elevation model (DEM) is supplied and `use_tobler = TRUE`,
movement cost is adjusted using Tobler’s hiking function, which increases
effective distance across steep slopes.

```r
# Create a synthetic DEM on the same grid for a lightweight vignette example
bnd_v <- terra::vect(boundary_sf)
dem <- terra::rast(ext = terra::ext(bnd_v), resolution = 20, crs = terra::crs(bnd_v))

xy <- terra::crds(dem, df = TRUE)
y0 <- (min(xy$y) + max(xy$y)) / 2
sigma <- (max(xy$y) - min(xy$y)) / 10
height <- 800
terra::values(dem) <- height * exp(-((xy$y - y0)^2) / (2 * sigma^2))

out_geo_dem <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "geodesic",
  dem_rast = dem,
  use_tobler = TRUE,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_geo_dem$polygons["generator_id"], main = "Geodesic + Tobler resistance", add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

# Comparing Euclidean and geodesic tessellations

Although both tessellations fully partition the domain, they differ in how influence propagates through concave boundaries and narrow corridors. The Euclidean tessellation allocates regions based on straight-line distance within the rasterised domain, whereas the geodesic tessellation respects the domain geometry and prevents influence across inaccessible areas.

These differences are reflected both visually and in the generator-level summary tables returned by the software.

```r
out_euc$summary
out_geo$summary
```

# Practical considerations

- Resolution: Smaller raster cell sizes improve boundary fidelity but increase computation time.

- Weights: Alternative weight transformations (e.g. square-root or power-law) can be supplied to reflect different assumptions about generator influence.

- Distance metric: Geodesic distance is recommended when domain geometry strongly constrains access or interaction.


# Summary

This vignette illustrates how weightedVoronoi can be used to generate reproducible, weighted spatial tessellations that respect complex boundaries and heterogeneous generator influence. The approach provides a flexible alternative to standard Voronoi methods in ecological and social–ecological applications.