---
title: "Weighted spatial tessellations in constrained domains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighted spatial tessellations in constrained domains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup}
library(sf)
library(terra)
library(weightedVoronoi)
```

# Introduction

Spatial allocation around point locations is a recurring requirement in ecology and social–ecological research, including the delineation of settlement influence zones, species territories, and management areas. Voronoi tessellations provide a simple and transparent framework for dividing space among generators, but standard implementations assume uniform influence and unconstrained Euclidean distance.

The weightedVoronoi package extends this framework by allowing (i) generator-specific weights, (ii) tessellations constrained to arbitrary polygonal domains using either Euclidean or geodesic distance, (iii) optional resistance- and terrain-informed geodesic allocation, and (iv) uncertainty-aware and temporal tessellation workflows.


## Example data

We first define a concave, irregular spatial domain and a set of generator points with heterogeneous weights.

```r
crs_use <- 32636  # projected CRS (metres)

make_irregular_boundary <- function(crs = crs_use) {
  # A simple concave polygon (fast, deterministic, vignette-safe)
  coords <- rbind(
    c(0, 0),
    c(1200, 0),
    c(1200, 900),
    c(800, 900),
    c(800, 500),
    c(400, 500),
    c(400, 900),
    c(0, 900),
    c(0, 0)
  )
  st_sf(geometry = st_sfc(st_polygon(list(coords))), crs = crs)
}

boundary_sf <- make_irregular_boundary()

set.seed(42)
pts <- st_sample(boundary_sf, size = 8, type = "random")

points_sf <- st_sf(
  village = paste0("V", seq_along(pts)),
  population = round(exp(rnorm(length(pts), log(300), 0.8))),
  geometry = pts,
  crs = crs_use
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Weighted Euclidean tessellation

We first generate a weighted tessellation using Euclidean distance. Generator influence is scaled using a logarithmic transformation of the population weights.

```r
out_euc <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "euclidean",
  # Optional alternative weight semantics:
  weight_model = "multiplicative",
  clip_to_boundary = TRUE,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_euc$polygons["generator_id"], add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Weighted geodesic tessellation

Geodesic tessellations compute shortest-path distances constrained to the domain.

```r
out_geo <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "geodesic",
  close_mask = TRUE,
  close_iters = 1,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_geo$polygons["generator_id"], add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

By default, geodesic allocation uses the `"classic"` engine, which computes one
accumulated-cost surface per generator. For additive isotropic geodesic
tessellations, a faster `"multisource"` engine is also available:

```r
out_geo_fast <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  distance = "geodesic",
  weight_model = "additive",
  geodesic_engine = "multisource",
  verbose = FALSE
)
```
The multisource engine is currently available for additive isotropic geodesic
tessellations (`weight_model = "additive"` and `anisotropy = "none"`).

## Custom resistance surfaces and barriers

In many ecological and social–ecological applications, effective distance is shaped not only by domain geometry but also by environmental resistance. Topography, land cover, and infrastructure can increase or block movement.

The recommended workflow is to build a resistance surface (optionally composed from multiple rasters), apply barriers, and then pass the result to weighted_voronoi_domain() via resistance_rast.

```r
# Build a template grid (aligned to the analysis)
template <- out_euc$allocation

# Two simple resistance layers (for demonstration)
R_land <- template; terra::values(R_land) <- 1
R_land[template == 1] <- 2  # small spatial pattern

R_risk <- template; terra::values(R_risk) <- 1
xy <- terra::xyFromCell(R_risk, 1:terra::ncell(R_risk))
band <- xy[,1] > 450 & xy[,1] < 550
vals <- terra::values(R_risk); vals[band] <- 10; terra::values(R_risk) <- vals

# Compose layers (multiplicative by default)
R <- compose_resistance(R_land, R_risk, template = template, method = "multiply")

# Add a river barrier (semi-permeable)
river <- st_sf(
  geometry = st_sfc(st_linestring(rbind(c(500, 0), c(500, 900)))),
  crs = st_crs(boundary_sf)
)

R2 <- add_barriers(R, river, permeability = "semi", cost_multiplier = 20, width = 20)

out_geo_res <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "geodesic",
  resistance_rast = R2,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_geo_res$polygons["generator_id"], add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Geodesic tessellations with isotropic elevation-dependent resistance

In many ecological and social–ecological applications, effective distance is
shaped not only by domain geometry but also by environmental resistance.
Topography, land cover, or infrastructure may increase the cost of movement
between locations. To accommodate this, `weightedVoronoi` allows geodesic
distance calculations to be modified by an isotropic resistance surface.

When a digital elevation model (DEM) is supplied and `use_tobler = TRUE`,
movement cost is adjusted using Tobler’s hiking function, which increases
effective distance across steep slopes.

```r
# Create a synthetic DEM on the same grid for a lightweight vignette example
bnd_v <- terra::vect(boundary_sf)
dem <- terra::rast(ext = terra::ext(bnd_v), resolution = 20, crs = terra::crs(bnd_v))

xy <- terra::crds(dem, df = TRUE)
y0 <- (min(xy$y) + max(xy$y)) / 2
sigma <- (max(xy$y) - min(xy$y)) / 10
height <- 800
terra::values(dem) <- height * exp(-((xy$y - y0)^2) / (2 * sigma^2))

out_geo_dem <- weighted_voronoi_domain(
  points_sf = points_sf,
  weight_col = "population",
  boundary_sf = boundary_sf,
  res = 20,
  weight_transform = log10,
  distance = "geodesic",
  dem_rast = dem,
  use_tobler = TRUE,
  verbose = FALSE
)

plot(st_geometry(boundary_sf), border = "black", lwd = 2)
plot(out_geo_dem$polygons["generator_id"], main = "Geodesic + Tobler resistance", add = TRUE)
plot(st_geometry(points_sf), add = TRUE, pch = 21, bg = "red")
```

## Terrain-anisotropic geodesic tessellations

The DEM-based geodesic workflow above uses isotropic resistance: movement cost
depends on slope magnitude, but not on movement direction. In some applications,
however, uphill and downhill movement should differ. For example, steep uphill
travel may be more costly than downhill travel, even across the same terrain.

`weightedVoronoi` therefore also supports a terrain-anisotropic geodesic mode,
in which movement between neighbouring raster cells becomes direction-dependent.
This requires a DEM and uses a directed transition graph internally.

Below, we use a simple synthetic eastward-rising DEM and compare isotropic and
terrain-anisotropic geodesic allocation.

```{r}
# Terrain-anisotropy example (self-contained)
crs_use2 <- 32636

boundary_sf2 <- st_sf(
  id = 1,
  geometry = st_sfc(
    st_polygon(list(rbind(
      c(0, 0),
      c(1200, 0),
      c(1200, 900),
      c(800, 900),
      c(800, 500),
      c(400, 500),
      c(400, 900),
      c(0, 900),
      c(0, 0)
    ))),
    crs = crs_use2
  )
)

dem_aniso <- terra::rast(
  ext = terra::ext(terra::vect(boundary_sf2)),
  resolution = 20,
  crs = terra::crs(terra::vect(boundary_sf2))
)

xy2 <- terra::crds(dem_aniso, df = TRUE)
terra::values(dem_aniso) <- xy2$x * 10

pts2 <- st_sf(
  village = c("A", "B"),
  population = c(1, 1),
  geometry = st_sfc(
    st_point(c(200, 450)),
    st_point(c(950, 450))
  ),
  crs = crs_use2
)

out_geo_iso2 <- weighted_voronoi_domain(
  points_sf = pts2,
  weight_col = "population",
  boundary_sf = boundary_sf2,
  res = 20,
  distance = "geodesic",
  dem_rast = dem_aniso,
  use_tobler = TRUE,
  anisotropy = "none",
  verbose = FALSE
)

out_geo_aniso2 <- weighted_voronoi_domain(
  points_sf = pts2,
  weight_col = "population",
  boundary_sf = boundary_sf2,
  res = 20,
  distance = "geodesic",
  dem_rast = dem_aniso,
  use_tobler = TRUE,
  anisotropy = "terrain",
  uphill_factor = 3,
  downhill_factor = 1.2,
  verbose = FALSE
)

par(mfrow = c(1, 2))
plot(st_geometry(boundary_sf2), border = "black", lwd = 2, main = "Isotropic DEM geodesic")
plot(out_geo_iso2$polygons["generator_id"], add = TRUE)
plot(st_geometry(pts2), add = TRUE, pch = 21, bg = "red")

plot(st_geometry(boundary_sf2), border = "black", lwd = 2, main = "Terrain-anisotropic geodesic")
plot(out_geo_aniso2$polygons["generator_id"], add = TRUE)
plot(st_geometry(pts2), add = TRUE, pch = 21, bg = "red")

par(mfrow = c(1, 1))
```

In this example, uphill and downhill movement are treated differently, so the
terrain-anisotropic tessellation can diverge from the isotropic DEM-based result.
The parameters `uphill_factor` and `downhill_factor` control the strength of this
directional asymmetry.

## Uncertainty-aware tessellations

In some applications, generator weights are not known exactly. To explore how
this uncertainty affects allocation, `weightedVoronoi` can repeat the
tessellation under stochastic perturbation of generator weights and summarise the
results as per-cell membership probabilities and entropy.

In the current implementation, uncertainty is applied to generator weights only.
For each simulation, weights are perturbed using a lognormal multiplicative
model, the tessellation is recomputed, and the allocation rasters are combined
into probability surfaces. Entropy is then used to summarise spatial
uncertainty: low entropy indicates stable assignment, whereas high entropy
indicates cells that switch more often among generators.

```{r}
crs_u <- "EPSG:32636"

boundary_u <- st_sf(
  geometry = st_sfc(st_polygon(list(rbind(
    c(0, 0),
    c(200, 0),
    c(200, 200),
    c(0, 200),
    c(0, 0)
  )))),
  crs = crs_u
)

points_u <- st_sf(
  population = c(0.01, 0.02),
  geometry = st_sfc(
    st_point(c(60, 100)),
    st_point(c(140, 100))
  ),
  crs = crs_u
)

out_u <- weighted_voronoi_uncertainty(
  points_sf = points_u,
  weight_col = "population",
  boundary_sf = boundary_u,
  n_sim = 100,
  weight_sd = 0.8,
  distance = "geodesic",
  geodesic_engine = "multisource",
  weight_model = "additive",
  verbose = FALSE,
  seed = 1
)

plot(out_u$entropy, main = "Entropy")
plot(terra::vect(points_u), add = TRUE, pch = 21, col = "black", bg = "red")
```

## Temporal tessellations

`weightedVoronoi` can also run tessellations across a sequence of time-specific
point datasets. This makes it possible to compare allocation through time while
allowing generator weights, locations, and resistance inputs to vary by time step.

```{r}
pts_t1 <- points_u
pts_t2 <- points_u
pts_t2$population <- c(0.02, 0.01)

out_time <- weighted_voronoi_time(
  points_list = list(time1 = pts_t1, time2 = pts_t2),
  weight_col = "population",
  boundary_sf = boundary_u,
  distance = "geodesic",
  geodesic_engine = "multisource",
  weight_model = "additive",
  verbose = FALSE
)

par(mfrow = c(1, 2))
plot(out_time$change_map_first_last, main = "Change map")
plot(out_time$persistence, main = "Persistence")
par(mfrow = c(1, 1))
```

In this example, `change_map_first_last` identifies cells whose allocation differs
between the first and last time step, while `persistence` highlights cells that
retain the same allocation across the full temporal sequence.

# Comparing Euclidean and geodesic tessellations

Although both tessellations fully partition the domain, they differ in how influence propagates through concave boundaries and narrow corridors. The Euclidean tessellation allocates regions based on straight-line distance within the rasterised domain, whereas the geodesic tessellation respects the domain geometry and prevents influence across inaccessible areas.

These differences are reflected both visually and in the generator-level summary tables returned by the software.

```r
out_euc$summary
out_geo$summary
```

# Practical considerations

- Resolution: Smaller raster cell sizes improve boundary fidelity but increase computation time.

- Weights: Alternative weight transformations (e.g. square-root or power-law) can be supplied to reflect different assumptions about generator influence.

- Distance metric: Geodesic distance is recommended when domain geometry strongly constrains access or interaction.

- Terrain anisotropy: Use `anisotropy = "terrain"` with `dem_rast` when uphill and downhill movement should differ.

- Uncertainty: If entropy is zero everywhere, the tessellation may simply be robust to the perturbation level used; stronger perturbations or a smaller spatial domain may be needed to reveal uncertainty.

- Temporal analysis: `weighted_voronoi_time()` supports time-specific point datasets and can return change and persistence maps.

# Summary

This vignette illustrates how weightedVoronoi can be used to generate reproducible, weighted spatial tessellations that respect complex boundaries and heterogeneous generator influence. The approach provides a flexible alternative to standard Voronoi methods in ecological and social–ecological applications.